Query Optimization for Dynamic Imputation

### 2017-05-15 paper/experiments.tex
paper: break up results section

\subsection{Results}\label{sec:results}

@@

\Cref{fig:runtimes} summarizes the performance results. The quality-optimized
queries are an order-of-magnitude faster than the baseline. We get another order-of-magnitude speedup when optimizing for performance.

@@

%\begin{figure}
%\includegraphics[width=\columnwidth]{figures/planning_times_imputedb.png}
%\caption{Planning times for each query}
%\label{fig:plantimes}
%\end{figure}
\subsubsection{Accuracy vs. Base-table Imputation}

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for \ProjectName{}'s query results compared to the baseline, using the reference chained-equation decision trees model.
This measures the error introduced by on-the-fly imputation as compared to full imputation on the base tables.
This comparison is the relevant one in practice --- an analyst would only be considering the trade-offs in imputation quality and time in the first place for dirty input data, for which ground truth is unavailable.

@@

\review{
We also experiment with mean-value imputation and hot
deck imputation (\Cref{sec:imputation}). In our implementation of mean-value imputation,
to facilitate non-blocking operation within the iterator
model, we estimate the column mean from base table histograms. Therefore,
imputations are identical no matter where the operator is placed in the query
plan. Moreover, runtimes are essentially identical regardless of the value of
$\alpha$ because the time cost of the imputation is relatively low no matter
how many tuples need to be imputed. In hot deck imputation, runtimes are also
similar across $\alpha$, because the cost of buffering
tuples and sampling at random is negligible compared to the rest of query
execution. \todo{be quantitative: In terms of accuracy, relative errors using hot deck are close to
zero; since the baseline, in this case, is hot deck imputation on the base
table, only skew in the distributions arising from filters and joins would lead
to larger errors. In some cases, skew does arise, and errors are on the same
magnitude as in the decision trees case. Indeed, \ProjectName{} yields larger
benefit when using relatively sophisticated, higher quality imputation
algorithms as opposed to simpler strategies that would even be quick to
implement on the base table.
}
}

We also calculate the number of tuples used to produce each aggregate output.
The count fraction columns in \Cref{table:smape} show the number of tuples in the aggregate for $\alpha = 0$ and $\alpha = 1$ as a fraction of the number of tuples used when running the query on the imputed base table.
This shows that when optimizing for performance, not quality, many tuples are dropped due to
insertion of $\lambda$ operators.
Even in cases where the SMAPE reduction from $\alpha = 1$ and $\alpha = 0$ is small
(\Cref{q2} and \Cref{q6}) the tuple count is significantly different.
In these cases, the aggregate value is not significantly impacted by the missing data. In
particular, if values are missing completely at random,
the aggregate should not be affected.
However, if the missing data is biased then the aggregate will have a significant error.
This highlights a challenge for a user handling data imputation traditionally: it is unclear if the missing data will have a large or small negative impact on their analysis until they have paid the cost of running it.
\review{
By using \ProjectName{} this cost can be lowered significantly. We can also trivially extend \ProjectName{} to warn users when the query chosen for execution has a high
$\textsc{Penalty}$ estimate, along with the number of tuples that have been dropped (after execution), so that situations with high potential for skewed results can be identified by the user.

\Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced by the planner for \Cref{q1} and \Cref{q6}. These are the resulting set of plans after pruning dominated plans throughout our algorithm
and encapsulate the possible trade-offs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
can be exposed to the user to help in guiding their workflow.
}

@@

Planning times for our benchmark queries are reasonable, with mean planning times across the
benchmark queries ranging from \planningtimelow{} to \planningtimehigh{}, and ranging from
\planningtimelowpct{} to \planningtimehighpct{}\% of total runtime. In 99\% of individual
cases, the optimizer returned a query plan within \planningtimepninetynine{}. The choice of
$\alpha$ has no impact on planning times, because query selection is done after collecting
the Pareto frontier.

\review{
However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
We explored the limits of our planning algorithm by constructing a series of queries with increasing number of joins over the CDC tables.
Planning times for queries involving 1--5 joins---a practical value for  real world exploratory queries---are less than 1 second.
We extended \ProjectName{} to support an approximation of Pareto sets to improve planning times for queries with 6 or more joins. Approximate Pareto sets
displayed a reduction in planning time linear with respect to the reduction in size of the final frontier. For queries with 6--8 joins, approximate sets achieved a reduction
in planning time between 14\% and 23\%, on average. Users can increase the level of approximation to reduce planning times further.
}

\subsubsection{Baseline Runtime}
In many real-world cases, applying imputation to an entire dataset is prohibitively expensive.
For example, if we run chained-equation decision trees imputation on the full ACS dataset, it completes in \acsbaseresultminutes{}.

In contrast, \ProjectName{} executes the query \lstinline{SELECT AVG(c0) FROM acs} over the ACS dataset (also using chained-equation decision trees imputation) in \acsimputedbzeroresult{} when optimizing for quality and \acsimputedboneresult{} when optimizing for runtime. 
\review{Given the runtime difference, we can run approximately 5,000 queries before taking more time than full imputation.
The performance difference comes from the fact that \ProjectName{} only needs to perform imputation on the single column required by the query.
Adding selection predicates would further reduce the query runtime by reducing the amount of required imputation.
An analyst could do the same, but tracking the required imputations would get increasingly complicated as the queries became more complex.
This highlights the benefit of using our system for early data exploration.}

>>>>

\review{
\subsection{Results}\label{sec:results}

@@

\textbf{Runtime vs. Base-table Imputation}\quad \Cref{fig:runtimes} summarizes the performance results. The quality-optimized
queries are an order-of-magnitude faster than the baseline. We get another order-of-magnitude speedup when optimizing for performance.

@@

%\begin{figure}
%\includegraphics[width=\columnwidth]{figures/planning_times_imputedb.png}
%\caption{Planning times for each query}
%\label{fig:plantimes}
%\end{figure}
\textbf{Accuracy vs. Base-table Imputation (AVG)}\quad \Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for \ProjectName{}'s query results compared to the baseline, using the reference chained-equation decision trees model.
This measures the error introduced by on-the-fly imputation as compared to full imputation on the base tables.
This comparison is the relevant one in practice --- an analyst would only be considering the trade-offs in imputation quality and time in the first place for dirty input data, for which ground truth is unavailable.

@@

\textbf{Accuracy vs. Base-table Imputation (COUNT)}\quad
We also calculate the number of tuples used to produce each aggregate output.
The count fraction columns in \Cref{table:smape} show the number of tuples in the aggregate for $\alpha = 0$ and $\alpha = 1$ as a fraction of the number of tuples used when running the query on the imputed base table.
This shows that when optimizing for performance, not quality, many tuples are dropped due to
insertion of $\delta$ operators.
Even in cases where the SMAPE reduction from $\alpha = 1$ and $\alpha = 0$ is small
(\Cref{q2} and \Cref{q6}) the tuple count is significantly different.
In these cases, the aggregate value is not significantly impacted by the missing data. In
particular, if values are missing completely at random,
the aggregate should not be affected.
However, if the missing data is biased then the aggregate will have a significant error.
This highlights a challenge for a user handling data imputation traditionally: it is unclear if the missing data will have a large or small negative impact on their analysis until they have paid the cost of running it.

By using \ProjectName{} this cost can be lowered significantly. We can also trivially extend \ProjectName{} to warn users when the query chosen for execution has a high
$\textsc{Penalty}$ estimate, along with the number of tuples that have been dropped (after execution), so that situations with high potential for skewed results can be identified by the user.

\textbf{Alternate Imputation Methods}\quad
We also experiment with mean-value imputation and hot
deck imputation (\Cref{sec:imputation}). In our implementation of mean-value imputation,
to facilitate non-blocking operation within the iterator
model, we estimate the column mean from base table histograms. Therefore,
imputations are identical no matter where the operator is placed in the query
plan. Moreover, runtimes are essentially identical regardless of the value of
$\alpha$ because the time cost of the imputation is relatively low independent of
how many tuples need to be imputed. In hot deck imputation, runtimes are also
similar across $\alpha$, because the cost of buffering
tuples and sampling at random is negligible compared to the rest of query
execution. \todo{be quantitative: In terms of accuracy, relative errors using hot deck are close to
zero; since the baseline, in this case, is hot deck imputation on the base
table, only skew in the distributions arising from filters and joins would lead
to larger errors. In some cases, skew does arise, and errors are on the same
magnitude as in the decision trees case. Indeed, \ProjectName{} yields larger
benefit when using relatively sophisticated, higher quality imputation
algorithms as opposed to simpler strategies that would even be quick to
implement on the base table.
}

\textbf{Pareto Frontiers}\quad
\Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced by the planner for \Cref{q1} and \Cref{q6}. These are the resulting set of plans after pruning dominated plans throughout our algorithm
and encapsulate the possible trade-offs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
can be exposed to the user to guide their workflow.

@@

\textbf{Planning Times}\quad
Planning times for our benchmark queries are reasonable, with mean planning times across the
benchmark queries ranging from \planningtimelow{} to \planningtimehigh{}, and ranging from
\planningtimelowpct{} to \planningtimehighpct{}\% of total runtime. In 99\% of individual
cases, the optimizer returned a query plan within \planningtimepninetynine{}. The choice of
$\alpha$ has no impact on planning times, because query selection is done after collecting
the Pareto frontier.

  \textbf{Approximate Pareto Sets}\quad
However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
We explored the limits of our planning algorithm by constructing a series of queries with increasing number of joins over the CDC tables.
Planning times for queries involving 1--5 joins---a practical value for  real world exploratory queries---are less than 1 second.
We extended \ProjectName{} to support an approximation of Pareto sets to improve planning times for queries with 6 or more joins. Approximate Pareto sets
displayed a reduction in planning time linear with respect to the reduction in size of the final frontier. For queries with 6--8 joins, approximate sets achieved a reduction
in planning time between 14\% and 23\%, on average. Users can increase the level of approximation to reduce planning times further.

\textbf{Imputation on Large Data}\quad
In many real-world cases, applying imputation to an entire dataset is prohibitively expensive.
For example, if we run chained-equation decision trees imputation on the full ACS dataset, it completes in \acsbaseresultminutes{}.

In contrast, \ProjectName{} executes the query \lstinline{SELECT AVG(c0) FROM acs} over the ACS dataset (also using chained-equation decision trees imputation) in \acsimputedbzeroresult{} when optimizing for quality and \acsimputedboneresult{} when optimizing for runtime. 
Given the runtime difference, we can run approximately 5,000 queries before taking more time than full imputation.
The performance difference comes from the fact that \ProjectName{} only needs to perform imputation on the single column required by the query.
Adding selection predicates would further reduce the query runtime by reducing the amount of required imputation.
An analyst could do the same, but tracking the required imputations would get increasingly complicated as the queries became more complex.
This highlights the benefit of using our system for early data exploration.
}
